---
title: "Troubleshooting & FAQ"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting & FAQ}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette covers common issues you may encounter when using `GiottoDB` and how to resolve them.

## Giotto functions do not work with GiottoDB objects

### Symptom

After loading `Giotto`, `GiottoClass`, `GiottoVisuals`, or the full Giotto Suite
after `GiottoDB`, calls such as `runPCA(gobject)`, `normalizeGiotto(gobject)`,
`calculateHVF(gobject)`, or `saveGiotto(gobject)` may use the standard Giotto
implementation instead of the GiottoDB implementation.

### Cause

Some Giotto functions are exported as regular functions rather than shared
generics. In R, packages loaded later appear earlier on the search path. Loading
`Giotto` after `GiottoDB` can therefore mask GiottoDB wrappers with functions
of the same name.

```{r, eval = FALSE}
library(GiottoDB)
library(Giotto)

# In this load order, runPCA() resolves to Giotto::runPCA().
find("runPCA")
```

This does not remove the GiottoDB implementation. It only changes which function
is found first when an unqualified function name is used.

### Solutions

Always load `GiottoDB` last when using unqualified function names:

```{r, eval = FALSE}
library(Giotto)
library(GiottoDB)

gobject <- normalizeGiotto(gobject)
gobject <- calculateHVF(gobject)
gobject <- runPCA(gobject)
saveGiotto(gobject, foldername = "my_giottodb_object")
```

Alternatively, call GiottoDB functions explicitly:

```{r, eval = FALSE}
gobject <- GiottoDB::normalizeGiotto(gobject)
gobject <- GiottoDB::calculateHVF(gobject)
gobject <- GiottoDB::runPCA(gobject)
GiottoDB::saveGiotto(gobject, foldername = "my_giottodb_object")
```

This is especially important for functions with GiottoDB-specific
database-backed behavior, including expression processing, PCA, saving and
loading, and selected visualization methods.

A longer-term solution is planned to define more shared generics upstream in Giotto or GiottoClass, with Giotto and GiottoDB each
registering methods for their supported object classes.

## Why does this happen?

GiottoDB extends Giotto workflows by adding methods and wrappers for
database-backed objects when needed. For functions that are not shared S3 or S4 generics in
the Giotto API, R cannot dispatch to GiottoDB methods if a later-loaded package
masks the GiottoDB wrapper.

## How can I check which function is being called?

Use `find()` or inspect the function environment:

```{r, eval = FALSE}
find("runPCA")
environmentName(environment(runPCA))
environmentName(environment(GiottoDB::runPCA))
```

If the unqualified function resolves to `Giotto`, `GiottoClass`, or
`GiottoVisuals`, use `GiottoDB::function_name()` or reload `GiottoDB` last.

## Getting help

If you encounter issues not covered here:

1. Visit the [GiottoDB issue page](https://github.com/giotto-suite/giottodb/issues).
2. Search for existing issues/errors that match your problem.
3. If you cannot find a solution, open a new issue and include a minimal reproducible example.

