---
title: "Tutorial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(GiottoDB)
```

## Introduction
- Link the Overview vignette.
- Demonstration of analyzing a MERFISH dataset with GiottoDB.

## Creating a GiottoDB Object
We will use the [GiottoData](https://github.com/drieslab/GiottoData) package to load a small spatial omics dataset (MERFISH).

```{r}
options("giotto.use_conda" = FALSE)
gobject <- GiottoData::loadGiottoMini(dataset = 'vizgen')

tmpfile = file.path(tempdir(), "test.db")
con = DBI::dbConnect(duckdb::duckdb(), tmpfile)
gobject_db <- as_giottodb(gobject, con = con)
```


```{r}

```

### Convenience function to convert existing Giotto objects into GiottoDB objects
Alternatively, we can use the `as_giottodb()` function to convert an existing Giotto object into a GiottoDB object.
```{r}
gobject_db <- as_giottodb(gobject)
```


## Loading MERFISH Data
```{r}

```


## Filtering 
```{r}

```


## Normalization
```{r}

```

## 
