---
title: "visium-hd-3p-ovarian"
vignette: >
  %\VignetteIndexEntry{visium-hd-3p-ovarian}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

# Introduction

Analysis pipeline for integrating alternative polyadenylation (APA) data from `scAPAtrap` with Visium HD 3' spatial transcriptomics data. 

The workflow includes:
1. Loading preprocessed APA peak data (PACdataset) generated from `process_scapatrap.R`
2. Integration of APA peaks with Space Ranger gene expression and spatial data using `GiottoSeq`
3. Giotto object creation combining APA, gene expression, RUD scores, and spatial coordinates

# Load Required Packages

```{r setup, message=FALSE, warning=FALSE}
#| label: setup
# required_packages <- c(
#   "giottodb" = "dbverse-org/giottodb@f9f077a",
#   "GiottoUtils" = "drieslab/GiottoUtils@f766cf3",
#   "GiottoClass" = "drieslab/GiottoClass@7e717ac4",
#   "Giotto" = "drieslab/Giotto@15c1078e",
#   "GiottoSeq" = "drieslab/GiottoSeq@c387816",
#   "dbMatrix" = "dbverse-org/dbmatrix-r@dev",
#   "dbSpatial" = "dbverse-org/dbspatial-r@dev"
# )

# # Install/update packages to exact commits
# if (!requireNamespace("pak", quietly = TRUE)) {
#   install.packages("pak")
# }
# pak::pak(required_packages, ask = FALSE)

if (!requireNamespace("pacman", quietly = TRUE)) {
    install.packages("pacman")
}

pacman::p_load(
    GiottoDB,      # dbverse-org/giottodb@f9f077a (main branch)
    GiottoUtils,   # drieslab/GiottoUtils@f766cf3 (dev branch)
    GiottoClass,   # drieslab/GiottoClass@7e717ac4 (db branch)
    Giotto,        # drieslab/Giotto@15c1078e (db branch)
    GiottoSeq,     # drieslab/GiottoSeq@c387816 (dev branch)
    dbMatrix,      # dbverse-org/dbmatrix-r (dev branch)
    dbSpatial,     # dbverse-org/dbspatial-r (dev branch)
    movAPA,
    Matrix,
    DBI,
    duckdb,
    dplyr,
    data.table,
)
```

# Dataset

TODO: Add a few words on the ovarian cancer 10X Genomics Visium HD 3' dataset. 

[ovarian.png](link)

# Load Preprocessed PACdataset

Load the preprocessed PACdataset generated by `process_scapatrap.R`.

```{r load-pacds}
pacds <- readRDS(
    '/Users/user/Documents/manuscript-olap-spatial-omics/fig2/PACds_filtered_vhd3p_oc.RDS'
)

summary(pacds)
```

# Load Outputs from 10X Genomics SpaceRanger 4.0 Pipeline

Below we are using the dbverse approach to load the outputs from the Space Ranger pipeline.

TODO: A convenience function to streamline this process.

## Setup Database Connection

```{r setup-database}
con <- DBI::dbConnect(
    duckdb::duckdb(),
    'vhd3p_integrated_analysis.db',
    read_only = FALSE
)
```

## Load Data Paths

Setup paths for data generated from Space Ranger outputs.

```{r load-data-paths}
data_dir <- '/Users/user/Documents/ovarian_cancer_compoohkie'
mtx_path <- file.path(data_dir, 'raw_feature_cell_matrix/matrix.mtx.gz')
mtx_barcodes_path <- file.path(data_dir, 'raw_feature_cell_matrix/barcodes.tsv.gz')
mtx_features_path <- file.path(data_dir, 'raw_feature_cell_matrix/features.tsv.gz')
```

## Load Barcode Mappings

The barcode mapping file links 2µm spatial bins to segmented cell IDs.

```{r load-barcode-mappings}
barcode_mappings_path <- file.path(data_dir, 'barcode_mappings.parquet')

barcode_map <- dplyr::tbl(
    con,
    dplyr::sql(paste0("SELECT * FROM '", barcode_mappings_path, "'"))
)

# Filter valid mappings and remove '-1' suffix artifact
barcode_map_cleaned <- barcode_map |>
    dplyr::filter(!is.na(cell_id) & !is.na(square_002um)) |>
    dplyr::mutate(
        square_002um_clean = regexp_replace(square_002um, '-1$', '')
    ) |>
    dplyr::compute(
        name = "barcode_map_cleaned",
        temporary = TRUE,
        overwrite = TRUE
    )
```

## Load Gene Expression Matrix

Load the gene expression matrix using dbMatrix for efficient database-backed operations.

```{r load-gene-expression}
# Read expression matrix as dbMatrix
dbmat <- dbMatrix::dbMatrix(
    value = mtx_path,
    class = "dbSparseMatrix",
    con = con,
    overwrite = TRUE,
    name = "dbMatrix",
    mtx_rowname_file_path = mtx_features_path,
    mtx_rowname_col_idx = 2,
    mtx_colname_file_path = mtx_barcodes_path,
    mtx_colname_col_idx = 1
)

# Filter nonzero values
# Note: this matches the behavior of Giotto::get10Xmatrix()
dbmat_rowsums <- rowSums(dbmat)
dbmat_rowsums_bool <- dbmat_rowsums != 0
dbmat_nonzero <- dbmat[dbmat_rowsums_bool, ]
```

## Load Spatial Coordinates

```{r load-spatial-coordinates}
tissue_positions_path <- file.path(
    data_dir,
    'tissue_positions.parquet'
)

# Creates VIEW in database of tissue_positions
tissue_positions <- dbSpatial::dbSpatial(
    value = tissue_positions_path,
    name = "tissue_positions",
    con = con,
    overwrite = TRUE
)

# Create `point` geometry in tissue_positions and save in db
tissue_positions <- tissue_positions[] |>
    dplyr::mutate(geom = st_point(pxl_row_in_fullres, pxl_col_in_fullres)) |>
    dplyr::compute(name = "tissue_positions_2", temporary = FALSE, overwrite = TRUE)
```

# Align RNA Expression Data

Add RNA expression data from Space Ranger using dbMatrix for efficient database-backed operations.

## Create RNA Expression + APA Data

The RNA expression matrix from Space Ranger already uses cell IDs as column names.

```{r create-rna-expression}
# RNA matrix is already at cell-level - use directly
rna_expr_obj <- Giotto::createExprObj(
    expression_data = dbmat_nonzero,
    name = "raw",
    spat_unit = "cell",
    feat_type = "rna",
    expression_matrix_class = "dbMatrix"
)

rna_cell_ids <- colnames(dbmat_nonzero)
cat("RNA cells:", length(rna_cell_ids), "\n")
```

Pool APA data from 2µm resolution to cell-level to match RNA resolution.

## Pool APA Data from 2µm to Cell Level

Use database operations to efficiently aggregate APA counts from 2µm bins to cells.

```{r pool-apa-to-cells}
# Get APA matrix from PACdataset and convert to dbMatrix
apa_counts <- pacds@counts

# Create dbMatrix from sparse matrix (loads into database efficiently)
apa_dbmat <- dbMatrix::dbMatrix(
    value = apa_counts,
    class = "dbSparseMatrix",
    con = con,
    name = "apa_counts_2um",
    overwrite = TRUE
)

# Convert ijx indices to named table (peak_name, barcode_2um, x)
# TODO: This should be an internal dbMatrix function like .dbMatrix_to_named_tbl()
peak_names <- rownames(apa_dbmat)
barcode_2um_names <- colnames(apa_dbmat)

# Create lookup tables for indices to names
peak_lookup <- data.table::data.table(
    i = seq_along(peak_names),
    peak_name = peak_names
)
barcode_lookup <- data.table::data.table(
    j = seq_along(barcode_2um_names),
    barcode_2um = barcode_2um_names
)

# Copy lookup tables to database
peak_lookup_tbl <- dplyr::copy_to(con, peak_lookup, "peak_lookup_temp", temporary = TRUE, overwrite = TRUE)
barcode_lookup_tbl <- dplyr::copy_to(con, barcode_lookup, "barcode_lookup_temp", temporary = TRUE, overwrite = TRUE)

# Join dbMatrix table with lookup tables to add dimension names
apa_long_named <- apa_dbmat[] |>
    dplyr::left_join(peak_lookup_tbl, by = "i") |>
    dplyr::left_join(barcode_lookup_tbl, by = "j") |>
    dplyr::select(peak_name, barcode_2um, x)

# Remove '-1' suffix from 2µm barcodes to match barcode_map_cleaned
apa_long_named <- apa_long_named |>
    dplyr::mutate(barcode_2um_clean = regexp_replace(barcode_2um, '-1$', ''))

# Join with barcode mappings to get cell_ids
apa_with_cellid <- apa_long_named |>
    dplyr::inner_join(
        dplyr::select(barcode_map_cleaned, square_002um_clean, cell_id),
        by = c("barcode_2um_clean" = "square_002um_clean")
    )

# Aggregate (pool) APA counts by peak × cell_id using database GROUP BY + SUM
# TODO: Benchmark: Is this SOTA for APA pooling? It's considerably faster than the method in BMILAB/movAPA.
apa_pooled <- apa_with_cellid |>
    dplyr::group_by(peak_name, cell_id) |>
    dplyr::summarise(x = sum(x, na.rm = TRUE), .groups = "drop") |>
    dplyr::filter(x > 0) |>  # Keep only non-zero values for sparsity
    dplyr::compute(name = "apa_pooled_temp", temporary = TRUE, overwrite = TRUE)

cat("Creating pooled APA dbMatrix from aggregated data...\n")
# Create dbMatrix from pooled long-format table
apa_dbmat_pooled <- dbMatrix::dbMatrix_from_tbl(
    tbl = apa_pooled,
    rownames_colName = "peak_name",
    colnames_colName = "cell_id",
    name = "dbMatrix_apa_pooled",
    overwrite = TRUE
)

# Create expression object from pooled APA dbMatrix
apa_expr_obj <- Giotto::createExprObj(
    expression_data = apa_dbmat_pooled,
    name = "raw",
    spat_unit = "cell",
    feat_type = "apa",
    expression_matrix_class = "dbMatrix"
)

apa_cell_ids <- colnames(apa_dbmat_pooled)
cat("APA pooling complete. Cells:", length(apa_cell_ids), "\n")
```

## Harmonize RNA and APA Cell IDs

Ensure both RNA and APA have the same set of cells.

```{r align-rna-apa}
# Find common cells between RNA and APA
common_cells <- intersect(rna_cell_ids, apa_cell_ids)

cat("\n=== Data Alignment ===\n")
cat("RNA cells:", length(rna_cell_ids), "\n")
cat("APA cells:", length(apa_cell_ids), "\n")
cat("Common cells:", length(common_cells), "\n")

# Subset both matrices to common cells if needed
if (length(common_cells) < length(rna_cell_ids) || length(common_cells) < length(apa_cell_ids)) {
    cat("Subsetting to common cells...\n")
    
    # Subset RNA
    dbmat_rna_aligned <- dbmat_nonzero[, common_cells]
    dbmat_rna_aligned[] <- dbmat_rna_aligned[] |>
        dplyr::compute(
            name = "dbMatrix_rna_aligned",
            temporary = TRUE,
            overwrite = TRUE
        )
    dbmat_rna_aligned@name <- "dbMatrix_rna_aligned"
    rna_expr_obj <- Giotto::createExprObj(
        expression_data = dbmat_rna_aligned,
        name = "raw",
        spat_unit = "cell",
        feat_type = "rna",
        expression_matrix_class = "dbMatrix"
    )
    
    # Subset APA
    apa_dbmat_aligned <- apa_dbmat_pooled[, common_cells]
    apa_dbmat_aligned[] <- apa_dbmat_aligned[] |>
        dplyr::compute(
            name = "dbMatrix_apa_aligned",
            temporary = TRUE,
            overwrite = TRUE
        )
    apa_dbmat_aligned@name <- "dbMatrix_apa_aligned"
    apa_expr_obj <- Giotto::createExprObj(
        expression_data = apa_dbmat_aligned,
        name = "raw",
        spat_unit = "cell",
        feat_type = "apa",
        expression_matrix_class = "dbMatrix"
    )
    
    rna_cell_ids <- common_cells
} else {
    cat("No subsetting needed, all cells are common.\n")
}
```

Map spatial coordinates to cell IDs and aggregate to cell centroids using dbSpatial.

```{r create-spatial-locations}
# Map tissue positions to cell IDs and create spatial coordinates
tissue_positions_cellid <- tissue_positions |>
    dplyr::left_join(
        barcode_map_cleaned,
        by = c("barcode" = "square_002um")
    ) |>
    dplyr::filter(!is.na(cell_id)) |>
    dplyr::transmute(
        cell_id,
        sdimx = pxl_row_in_fullres,
        sdimy = pxl_col_in_fullres
    )

# Filter to cells in RNA data and aggregate to cell centroids
# Use local() to avoid DuckDB translation error
spat_locs_cellid <- tissue_positions_cellid |>
    dplyr::filter(cell_id %in% local(rna_cell_ids)) |>
    dplyr::group_by(cell_id) |>
    dplyr::summarise(
        sdimx = mean(sdimx, na.rm = TRUE),
        sdimy = mean(sdimy, na.rm = TRUE),
        .groups = 'drop'
    ) |>
    dplyr::compute(name = "spat_locs_cellid", temporary = TRUE)

# Collect aggregated cell centroids and prepare for Giotto
spat_locs_dt <- spat_locs_cellid |>
    dplyr::collect() |>
    data.table::as.data.table()

# Ensure proper column naming for Giotto (cell_ID, not cell_id)
data.table::setnames(spat_locs_dt, "cell_id", "cell_ID")

cat("\n=== Spatial Locations Created ===\n")
cat("RNA cells:", length(rna_cell_ids), "\n")
cat("Spatial locations:", nrow(spat_locs_dt), "\n")

# Create spatial locations object
spat_locs_obj <- Giotto::createSpatLocsObj(
    coordinates = spat_locs_dt,
    spat_unit = "cell",
    name = "raw"
)
```

# Create Integrated Giotto Object

```{r create-base-giotto}
# Create empty Giotto object
gobject <- Giotto::createGiottoObject() |> suppressWarnings()
```

## Add Expression and Spatial Locations to Giotto Object

Add both RNA and APA expression objects and spatial locations to the Giotto object.

```{r add-expression-spatial}
# Add RNA expression to the Giotto object
gobject <- Giotto::setExpression(
    gobject = gobject,
    x = rna_expr_obj,
    verbose = TRUE
)

# Add APA expression to the Giotto object
gobject <- Giotto::setExpression(
    gobject = gobject,
    x = apa_expr_obj,
    name = "apa",
    verbose = TRUE
)

# Add spatial locations to the Giotto object
gobject <- Giotto::setSpatialLocations(
    gobject = gobject,
    x = spat_locs_obj,
    verbose = TRUE
)
```

Placeholder to save in-memory gobject.
```{r add-expression-spatial}

dgc <- as.matrix(rna_expr_obj[], sparse = TRUE, names = TRUE)
rna_expr_obj@exprMat <- dgc

# Add RNA expression to the Giotto object
gobject2 <- Giotto::setExpression(
    gobject = gobject,
    x = rna_expr_obj,
    verbose = TRUE
)

dgc <- as.matrix(apa_expr_obj[], sparse = TRUE, names = TRUE)
apa_expr_obj@exprMat <- dgc

# Add APA expression to the Giotto object
gobject2 <- Giotto::setExpression(
    gobject = gobject2,
    x = apa_expr_obj,
    name = "apa",
    verbose = TRUE
)

# Add spatial locations to the Giotto object
gobject2 <- Giotto::setSpatialLocations(
    gobject = gobject2,
    x = spat_locs_obj,
    verbose = TRUE
)
```

## Calculate RUD (Relative Usage of Distal sites) Scores

Compute alternative polyadenylation switching metrics using movAPA's RUD index.

```{r calculate-rud}
# Step 1: Reconstruct PACdataset from pooled APA expression in Giotto object
# We need the original pacds for annotation information
pacds_original <- readRDS(
    '/Users/user/Documents/manuscript-olap-spatial-omics/fig2/PACds_filtered_vhd3p_oc.RDS'
)

# Extract pooled APA counts from Giotto object (already at cell-level)
apa_pooled_matrix <- Giotto::getExpression(
    gobject = gobject2,
    spat_unit = "cell",
    feat_type = "apa",
    name = "raw",
    output = "matrix"
)

cat("Pooled APA matrix dimensions:", dim(apa_pooled_matrix), "\n")

# Create new colData for pooled cells
pooled_coldata <- data.frame(
    cell_id = colnames(apa_pooled_matrix),
    row.names = colnames(apa_pooled_matrix)
)

# Create new PACdataset with pooled counts at cell resolution
pacds_pooled <- new(
    "PACdataset",
    counts = apa_pooled_matrix,
    colData = pooled_coldata,
    anno = pacds_original@anno,  # Reuse original peak annotations
    supp = pacds_original@supp   # Reuse original supplemental data
)

cat("Created pooled PACdataset with dimensions:", dim(pacds_pooled@counts), "\n")

# Step 2: Examine P/D distance distribution to guide parameter selection
nearby_dist_summary <- movAPA::getNearbyPAdist(pacds = pacds_pooled)
print(nearby_dist_summary)

# Based on the distribution:
# - minDist: Set just above minimum (e.g., 50 bp)
# - maxDist: Set well above mean to capture most sites (e.g., 30000 bp for human)
# - minRatio: Start permissive (0.01 = 1%) for discovery; increase to 0.05 if too many NaN

# Step 3: Identify 3'UTR proximal and distal APA site pairs
# This filters for gene-level P/D pairs within specified distance constraints
pacds_pd <- movAPA::get3UTRAPApd(
    pacds_pooled,
    minDist = 50,       # Minimum distance (just above min observed)
    maxDist = 30000,    # Maximum distance (well above mean, captures most sites)
    minRatio = 0.01,    # Keep sites with ≥1% relative expression (permissive)
    fixDistal = FALSE,  # Allow flexible distal site selection (recommended)
    addCols = 'pd'      # Add P/D annotation columns
)

cat("P/D filtered PACdataset dimensions:", dim(pacds_pd@counts), "\n")

# Step 4: Calculate smartRUD index with weights
# RUD = Relative Usage of Distal sites
#   - RUD score: 0 = only proximal (short 3'UTR), 1 = only distal (long 3'UTR)
#   - smartRUD: Weighted version that down-weights lowly-expressed genes
#   - Formula: RUD = Distal / (Proximal + Distal)
#
# Parameters:
#   - sRUD.oweight = TRUE: Return both RUD scores AND weights (P+D totals)
#   - clearPAT: Minimum P+D count threshold
#       * 0 = keep all (permissive, more NaN)
#       * 2 = remove P+D ≤ 2 (stringent, fewer NaN, recommended)
#       * 5 = remove P+D ≤ 5 (very stringent, high-confidence only)
rud_results <- movAPA::movAPAindex(
    pacds_pd,
    method = "smartRUD",
    sRUD.oweight = TRUE,  # Output both RUD values and weights (P+D totals)
    clearPAT = 0          # Keep all non-zero counts (permissive; try 2 if too many NaN)
)

# Extract RUD and weight matrices
rud_matrix <- rud_results$rud
weight_matrix <- rud_results$weight

# Display RUD distribution
print(summary(as.vector(rud_matrix[!is.na(rud_matrix)])))

# Step 4: Add RUD and weight matrices to Giotto object as separate expression layers
rud_expr_obj <- Giotto::createExprObj(
    expression_data = rud_matrix,
    name = "rud",
    spat_unit = "cell",
    feat_type = "apa"
)

weight_expr_obj <- Giotto::createExprObj(
    expression_data = weight_matrix,
    name = "weight",
    spat_unit = "cell",
    feat_type = "apa"
)

gobject2 <- Giotto::setExpression(
    gobject = gobject2,
    x = rud_expr_obj,
    verbose = TRUE
)

gobject2 <- Giotto::setExpression(
    gobject = gobject2,
    x = weight_expr_obj,
    verbose = TRUE
)

cat("\nRUD and weight matrices added to Giotto object\n")
```

## Save Integrated Giotto Object

Save in memory gobject
```{r save-giotto}
saveGiotto(
    gobject = gobject2,
    method = "qs",
    foldername = "giotto_memory_integrated_apa_rna",
    overwrite = TRUE
)

gobject2 <- loadGiotto("giotto_memory_integrated_apa_rna")
```

```{r save-giotto}
# Save integrated Giotto object using GiottoDB s3 method
gobject_db <- as_giottodb(gobject2, con = con)

saveGiotto(
    gobject = gobject_db,
    method = "qs",
    foldername = "gdb_vhd3p_integrated_apa_rna",
    overwrite = TRUE
)

gobject <- loadGiotto(
    path_to_folder = "gdb_vhd3p_integrated_apa_rna",
)
```
# Session Info

```{r session-info}
sessionInfo()
```

# Cleanup

```{r cleanup}
DBI::dbDisconnect(con, shutdown = TRUE)
```
